---
title: "Group Project_4830"
output: html_document
date: "2024-11-11"
---

```{r}
#Load libraries
library(tidyverse)
library(kableExtra)
library(rsample)
library(recipes)
library(parsnip)
library(yardstick)
library(viridisLite)
library(GGally)
```

```{r}
Heart<-read.csv("processed.cleveland.data")

#View(Heart)
```

```{r}
names <- c("Age",
           "Sex",
           "Chest_Pain_Type",
           "Resting_Blood_Pressure",
           "Serum_Cholesterol",
           "Fasting_Blood_Sugar",
           "Resting_ECG",
           "Max_Heart_Rate_Achieved",
           "Exercise_Induced_Angina",
           "ST_Depression_Exercise",
           "Peak_Exercise_ST_Segment",
           "Num_Major_Vessels_Flouro",
           "Thalassemia",
           "Diagnosis_Heart_Disease")
```

```{r}
head(Heart)
```

```{r}
table(Heart$Thalassemia)
```

```{r}
colnames(Heart) <- names
View(Heart)
```

```{r}
nrow(Heart)
```

```{r}
missing_values <- sapply(Heart, function(x) sum(is.na(x)))

print(missing_values)
```

```{r}
Heart[Heart == '?'] <- NA
View(Heart)
```

```{r}
missing_values <- sapply(Heart, function(x) sum(is.na(x)))

print(missing_values)
```

```{r}
Heart <- na.omit(Heart)
nrow(Heart)
```

```{r}

#Determine the number of values in each level of dependent variable
Heart %>% 
  drop_na() %>%
  group_by(Diagnosis_Heart_Disease) %>%
  count() %>% 
  ungroup() %>%
  kable(align = rep("c", 2)) %>% kable_styling("full_width" = F)
```

Since any value above 0 in ‘Diagnosis_Heart_Disease’ (column 14) indicates the presence of heart disease, we can lump all levels \> 0 together so the classification predictions are binary – Yes or No (1 or 0).

```{r}
Heart$Diagnosis_Heart_Disease <- ifelse(Heart$Diagnosis_Heart_Disease > 0, 1, 0)

head(Heart)
```

```{r}
#table(Heart)
```

The categorical variables have different levels, mapping the same into a categorical data subset.

```{r}

library(dplyr)
categorical_data <- Heart %>% 
  mutate(Sex = if_else(Sex == 1, "MALE", "FEMALE"),
         Exercise_Induced_Angina = if_else(Exercise_Induced_Angina == 1, "YES" ,"NO"),
         Chest_Pain_Type = if_else(Chest_Pain_Type == 1, "TYPICAL ANGINA",
                      if_else(Chest_Pain_Type == 2, "ATYPICAL ANGINA",
                      if_else(Chest_Pain_Type == 3, "NON-ANGINAL PAIN","ASYMPTOMATIC"))),
         Resting_ECG = if_else(Resting_ECG == 0, "NORMAL",
                           if_else(Resting_ECG == 1, "ABNORMALITY", "PROBABLE OR DEFINITE")),
         Num_Major_Vessels_Flouro = as.factor(Num_Major_Vessels_Flouro),
         Thalassemia = as.factor(Thalassemia),
         Diagnosis_Heart_Disease = if_else(Diagnosis_Heart_Disease == 1, "YES", "NO")
         ) %>% 
  mutate_if(is.character, as.factor) %>% 
  dplyr::select(Diagnosis_Heart_Disease, Sex, Exercise_Induced_Angina, Chest_Pain_Type, Resting_ECG, Num_Major_Vessels_Flouro, Thalassemia)
```

```{r}
summary(categorical_data)
```
#### Pie chart for gender 

```{r}
gender_summary <- as.data.frame(table(categorical_data$Sex))
colnames(gender_summary) <- c("Gender", "Count")
# Calculate percentages
gender_summary$Percentage <- round(gender_summary$Count / sum(gender_summary$Count) * 100, 1)

gender_summary = as.data.frame(gender_summary)
ggplot(gender_summary, aes(x = "", y = Count, fill = Gender)) +
  geom_bar(stat = "identity", width = 1) + # Bar chart for pie slices
  coord_polar(theta = "y" ,start = 0, direction = -1) + # Convert to pie chart
  geom_text(
    aes(
      y = cumsum(Count) - Count / 2, 
      label = paste0(Percentage, "%"),
      
    ),
    color = "white",
    position = position_stack(vjust = 0.5)
  ) +
  labs(
    title = "Gender Distribution",
    x = NULL,
    y = NULL
  ) +
  theme_void() + # Clean chart
  theme(legend.position = "right")
```
Since the data for male is overrepresented the model will have a better predictive power for that.

#### Pie chart for Heart Diagnosis

```{r}
diagnosis_summary <- as.data.frame(table(categorical_data$Diagnosis_Heart_Disease))
colnames(diagnosis_summary) <- c("Diagnosis", "Count")

# Calculate percentages
diagnosis_summary$Percentage <- round(diagnosis_summary$Count / sum(diagnosis_summary$Count) * 100, 1)

ggplot(diagnosis_summary, aes(x = "", y = Count, fill = Diagnosis)) +
  geom_bar(stat = "identity", width = 1) + # Bar chart for pie slices
  coord_polar(theta = "y", start = 0, direction = -1) + # Convert to pie chart
  geom_text(
    aes(
      y = cumsum(Count) - Count / 2, 
      label = paste0(Percentage, "%")
    ),
    color = "white",
    position = position_stack(vjust = 0.5)
  ) +
  labs(
    title = "Heart Diagnosis Distribution",
    x = NULL,
    y = NULL
  ) +
  theme_void() + # Clean chart
  theme(legend.position = "right")
```
Class imbalance therefore we can expect reduced sensitivity for detecting heart disease cases (false negatives).

```{r}
ggplot(categorical_data, aes(x = "", fill = Sex)) +
  geom_bar(stat= "count", width = 1) +
  coord_polar("y", start = 0, direction = -1) +
  geom_text(aes(label = scales::percent(..count.. / sum(..count..))), stat = "count", position = position_stack(vjust = .5)) +
  
   labs(
    title = "Gender Distribution",
    x = NULL,
    y = NULL
  )
  theme_void()
```


```{r}

ggplot(categorical_data, aes(x = "", fill = Diagnosis_Heart_Disease)) +
  geom_bar(stat= "count", width = 1) +
  coord_polar("y", start = 0, direction = -1) +
  geom_text(aes(label = scales::percent(..count.. / sum(..count..))), stat = "count", position = position_stack(vjust = .5)) +
  
   labs(
    title = "Heart Diagnosis Distribution",
    x = NULL,
    y = NULL
  )
  theme_void()
```

####Distribution of the categorical data.
```{r}
for (pair in colnames(categorical_data)) {
  
  tbl <- with(categorical_data, table(categorical_data[['Sex']], categorical_data[[pair]]))
  
  barplot(tbl, beside = TRUE, 
          col = c("skyblue", "salmon"), 
          main = paste("Comparison of", pair, "by", 'Sex'), 
          xlab = 'Sex', 
          ylab = "Count")
  
  legend("topright", legend = c('Female','Male'), fill = c("skyblue", "salmon"))
  }
```

In the dataset, the number of males are more than female around 200 patients are male and around 100 is only female.

```{r}
for (pair in colnames(categorical_data)) {
  
  tbl <- with(categorical_data, table(categorical_data[['Diagnosis_Heart_Disease']], categorical_data[[pair]]))
  
  barplot(tbl, beside = TRUE, 
          col = c("skyblue", "salmon"), 
          main = paste("Comparison of", pair, "by", 'Diagnosis'), 
          xlab = pair, 
          ylab = "Count", cex.names = 0.8)
  
  legend("topright", legend = c('No','Yes'), fill = c("skyblue", "salmon"))
  }
```

#### Density plots of the numerical variables.

```{r}
# Loop through each numerical variable for density plots
numerical_var = c('ST_Depression_Exercise','Age','Max_Heart_Rate_Achieved','Serum_Cholesterol','Resting_Blood_Pressure')
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = var, color = "factor(Diagnosis_Heart_Disease)")) +
    geom_density(alpha = 0.6, size = 1) +
    labs(title = paste("Density of", var, "by Heart Disease Diagnosis"),
         x = var,
         y = "Density",
         color = "Heart Disease") +
    theme_minimal()
    print(p)
}
```

#### Boxplot

```{r}
numerical_data = Heart[numerical_var]
boxplot(numerical_data, cex.axis=0.5)
```

```{r}
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = as.factor(categorical_data$Sex), y = var, fill = as.factor(Heart$Sex))) +
    geom_boxplot(outlier.color = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var, "by Gender"),
         x = "Gender",
         y = var) +
    theme_minimal()
    print(p)
}

```

```{r}
hist(numerical_data$Age)
```

```{r}
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = as.factor(categorical_data$Diagnosis_Heart_Disease), y = var, fill = as.factor(Heart$Diagnosis_Heart_Disease))) +
    geom_boxplot(outlier.color = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var, "by Heart Disease Diagnosis"),
         x = "Heart Diagnosis",
         y = var) +
    theme_minimal()
    print(p)
}
```

```{r}
for (var in numerical_var) {
  p <- ggplot(Heart, aes_string(x = as.factor(categorical_data$Chest_Pain_Type), y = var, fill = as.factor(categorical_data$Chest_Pain_Type))) +
    geom_boxplot(outlier.color = "red", outlier.shape = 16) +
    labs(title = paste("Boxplot of", var, "by Chest_Pain_Type"),
         x = "Chest_Pain_Type",
         y = var) +
    theme_minimal() + 
    theme(axis.text.x = element_text(face="bold", color="#993333", 
                           size=8, angle=45))
    print(p)
}
```

Scatterplot

```{r}
pairs(numerical_data)
```

#### Correlation

```{r}
library(corrplot)

#Using the cor function to find the correlation between the variables
cor_data = cor(numerical_data)
 
print("Correlation matrix")
print(cor_data)
```

```{r}
#correlation matrix 
corrplot(cor_data, method="number", cl.cex = 0.5, tl.col = "blue", tl.cex = 0.5)
```

##### Getting the numerical variables from the data + scaling

```{r}
heart_numerical <- scale(numerical_data)
```

### PCA

```{r}
heart.pca <- prcomp(heart_numerical)

#Eigen vectors or weights

heart.pca
```

```{r}
library(factoextra)
fviz_eig(heart.pca)  ## use the function on the pca object
```

Screeplot tells 2 principal components.

```{r}
library(factoextra)
fviz_pca_biplot(heart.pca, repel = TRUE,
                col.var = "steelblue", # Variables color
                col.ind = "coral")
```

```{r}
fviz_pca_var(heart.pca, col.var = "blue")
```

```{r}
#pca loadings
pca_loadings <- heart.pca$rotation
head(pca_loadings)
```

```{r}
#PCA scores
pca_scores <- heart.pca$x
head(pca_scores)

```

```{r}
#eigen values
eval <- (heart.pca$sdev)^2
eval
```

Kaiser Criterion

```{r}
sum(eval > mean(eval))
```

```{r}
percent_val <- eval/sum(eval)*100

percent_val
```

```{r}
cumsum(percent_val)
```

```{r}
cumulative_variance <- cumsum(percent_val)

# Create a data frame for the plot
df_variance <- data.frame(
  Component = 1:length(cumulative_variance),
  CumulativeVariance = cumulative_variance
)

# Plot cumulative percentage variance against component index
ggplot(df_variance, aes(x = Component, y = CumulativeVariance)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  labs(title = "Cumulative Percentage Variance vs Principal Component Index",
       x = "Principal Component Index",
       y = "Cumulative Percentage Variance (%)") +
  theme_minimal()
```

Cumsum method / 80-90% rule tells 4 principal components

```{r}
sub_dat <- as.data.frame(heart.pca$x[, 1:2])

plot(sub_dat$PC1, sub_dat$PC2, col = "blue",
     xlab = "PC1", ylab = "PC2", pch =18
)
```

```{r}
table(Heart$Diagnosis_Heart_Disease)
```

```{r}
pca_df <- data.frame(heart.pca$x[,1:2], heart = Heart$Diagnosis_Heart_Disease)

ggplot(pca_df, aes(PC1, PC2, color = heart)) + 
  geom_point(size = 3) + 
  labs(title = "PCA - Heart_Diagnosis", x = "PC1", y = "PC2") +
  theme_minimal()
```

```{r}
pca_df2 <- data.frame(heart.pca$x[,4:5], heart = Heart$Diagnosis_Heart_Disease)
ggplot(pca_df2, aes(PC4, PC5, color = heart)) + 
  geom_point(size = 3) + 
  labs(title = "PCA - Heart_Diagnosis", x = "PC1", y = "PC2") +
  theme_minimal()
```

```{r}
library(scatterplot3d)

# Assuming pca_data contains PC1, PC2, PC3 from PCA
pca3_data <- as.data.frame(heart.pca$x[, 1:3])  # Extract first 3 PCs
pca3_data$Diagnosis <- as.factor(Heart$Diagnosis_Heart_Disease)  # Add target variable

# Create the 3D scatterplot
scatterplot3d(pca_data$PC1, pca_data$PC2, pca_data$PC3,
              color = ifelse(pca_data$Diagnosis == 1, "red", "blue"),
              pch = 19,            # Solid circles
              xlab = "PC1",        # Label for X-axis
              ylab = "PC2",        # Label for Y-axis
              zlab = "PC3",        # Label for Z-axis
              main = "3D Scatterplot of PC1, PC2, PC3")
legend = c('Diagnosed','Not Diagnosed')

```

```{r}
library(scatterplot3d)

# Assuming pca_data contains PC1, PC2, PC3 from PCA
pca3_data <- as.data.frame(heart.pca$x[, 1:4])  # Extract first 3 PCs
pca3_data$Diagnosis <- as.factor(Heart$Diagnosis_Heart_Disease)  # Add target variable

# Create the 3D scatterplot
scatterplot3d(pca_data$PC1, pca_data$PC2, pca_data$PC4,
              color = ifelse(pca_data$Diagnosis == 1, "red", "blue"),
              pch = 19,            # Solid circles
              xlab = "PC1",        # Label for X-axis
              ylab = "PC2",        # Label for Y-axis
              zlab = "PC4",        # Label for Z-axis
              main = "3D Scatterplot of PC1, PC2, PC3")
legend = c('Diagnosed','Not Diagnosed')
```

```{r}
library(GGally)
#pca_data <- as.data.frame(pca_result$x)
#pca_data <- data.frame(heart.pca$x, heart = Heart$Diagnosis_Heart_Disease)
pca_data <- as.data.frame(heart.pca$x[, 1:5])

# Add the target variable to the PCA-transformed dataset
pca_data$Diagnosis <- Heart$Diagnosis_Heart_Disease  # Replace 'target' with the actual target variable

# Parallel coordinate plot
ggparcoord(data = pca_data, 
           columns = 1:5,         # Select PC1 to PC5
           groupColumn = "Diagnosis",
           scale = "globalminmax", 
           title = "Parallel Coordinate Plot of Principal Components",
           alphaLines = 0.5) +
  
  theme_minimal()

```

#### PCA prediction

```{r}
idx <- floor(0.8*nrow(Heart)) ### how many points to subset 
#idx
train_idx <- sample(1:nrow(Heart), idx) ##subset index for training set
```

```{r}
numerical_data$Diagnosis_Heart_Disease = Heart$Diagnosis_Heart_Disease
  
heart_train <- numerical_data[train_idx, ] ##subset rows for the train set 
heart_test <- numerical_data[-train_idx,  ] ##subset rows for the test set
heart_train
```

```{r}
heart_subset <- heart_train[,-6] ### Working on the training data set
heart_sub.pca <- prcomp(heart_subset, scale=TRUE)

summary(life_sub.pca)
```

```{r}
####transform the training variables into pca components using the fitted pca models
transform_train <- as.data.frame(predict(heart_sub.pca, heart_train))
head(transform_train)
```

```{r}
transform_train$Diagnosis_Heart_Disease = heart_train$Diagnosis_Heart_Disease
#transform_test = as.data.frame(transform_test)
head(transform_train)
```

```{r}
transform_test <- as.data.frame(predict(heart_sub.pca, heart_test)) ##transform the test variables into pca components using the fitted pca models

head(transform_test)
```

```{r}
transform_test$Diagnosis_Heart_Disease = heart_test$Diagnosis_Heart_Disease
#transform_test = as.data.frame(transform_test)
head(transform_test)
```

#### Creating the logistic regression model with the 2 principal components

```{r}
classifier = glm(formula = Diagnosis_Heart_Disease ~ .,
                 family = binomial,
                 data = transform_train[,c('PC1','PC2', 'PC3', 'PC4','Diagnosis_Heart_Disease')])
summary(classifier)
```

```{r}
prob_pred = predict(classifier, type = 'response', newdata = transform_test[-6])
y_pred = ifelse(prob_pred > 0.5, 1, 0)
```

```{r}
cm = table(transform_test[, 6], y_pred)
cm
```

```{r}
# Install and load caret package
#install.packages("caret")
library(caret)

# Convert to factors for compatibility
actual <- factor(heart_test$Diagnosis_Heart_Disease, levels = c(1, 0))  # Set levels explicitly
predicted <- factor((y_pred), levels = c(1, 0))

# Create confusion matrix
conf_matrix <- confusionMatrix(predicted, actual, positive = "1")

# Print confusion matrix
print(conf_matrix)

```

```{r}
library(pROC)

# Compute ROC curve
roc_curve <- roc(heart_test$Diagnosis_Heart_Disease, prob_pred)

# Plot the ROC curve
plot(roc_curve, main = "ROC Curve", legacy.axes = TRUE)
auc_value <- auc(roc_curve)
cat("AUC:", auc_value, "\n")

```
